\vspace{-2.25cm} 眼科疾病诊断的多模态异常感知模型

### 2024-10-26 Paper-Translated/Paper.tex
Translate paper.

\documentclass{article}
\usepackage{ctex}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{amsmath}
\usepackage{caption}
\RequirePackage{longtable,multirow,array}
\RequirePackage{booktabs}
\usepackage{booktabs}
\usepackage{multicol}
\usepackage{longtable}
\usepackage{xurl}
\usepackage{booktabs}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{enumitem}
\usepackage[style=ieee,sorting=nyt]{biblatex}
\assignrefcontextentries[]{*}
\addbibresource{References.bib}
\geometry{left=1in, right=1in, top=1in, bottom=1in}
\hypersetup{hidelinks}
\hypersetup{bookmarksnumbered=true,
	bookmarksopen=true,
	colorlinks=true,
	linkcolor=black,
	citecolor=blue,
	urlcolor=black}
\usepackage{tikz}
\usepackage{pgfplotstable}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\captionsetup{justification=centering}
\captionsetup[figure]{name={Fig.},labelsep=space,labelfont=bf}
\captionsetup[table]{name={Table},labelsep=space,labelfont=bf}
\CTEXoptions[today=old, contentsname=Contents]
\newcommand{\authyear}[1]{\citeauthor{#1} (\citeyear{#1})}
\setmainfont{Arial}

@@

\begin{document}
	\title{\vspace{-2.25cm} Multimodal Abnormity-Aware Model\\for Ocular Disease Diagnosis}
	\author{Siqi Pan}
	\date{}
	\maketitle

	\vspace{-0.2cm}
	\hrule
	\vspace{-0.1cm}
	\section*{Abstract}
	\hrule
	\vspace{0.4cm}
	We devise a novel model, the Multimodal Abnormity-Aware Model (MAAM), for assisting ocular disease diagnosis. The MAAM enables input of 2 kinds of ophthalmological images, namely optical coherence tomography (OCT) and color fundus images, and simulates the decision-making process by deducing the disease from abnormities. The model is composed of 2 submodels, the Abnormity Models and the Diagnosis Model, where the Abnormity Models identify 11 OCT abnormities and 8 fundus abnormities, and the Diagnosis Model identifies 12 diseases by integrating features from the former model. In MAAM, we introduce the fusion mechanism, the concept of codominant abnormities and diseases, and the Abnormity-to-Disease Criteria.  The model breaks the black box of complex neural network and expose more interpretable information for the reference of ophthalmologists. And it gets the overall accuracy of 89\% and thus the feasibility is verified.
	\vspace{0.4cm}
	\hrule
	\vspace{1cm}

	\begin{multicols}{2}
	\section{Background}
	Ocular diseases can be diagnosed through various methods, including using optical coherence tomography (OCT) and color fundus images. Ophthalmologists usually identify ocular abnormalities to deduce the disease. Traditionally, the diagnosis primarily depends on the professional experience and knowledge of the ophthalmologists, which may result in high misdiagnosis rate and under-utilization of medical data. With the widespread application of artificial intelligence (AI), deep learning (DL) has made great contributions in providing support to patients in remote areas by sharing expert knowledge \autocite{Ichhpujani_Thakur_2021}. By leveraging DL, researchers have developed auxiliary diagnosis programs to help ophthalmologists in the process. Many studies use convolutional neural networks (CNN) to analyze ocular images. Some commonly used CNNs are VGG, ResNet, and Inception \autocite{daich2023artificial}. And for segmenting images and finding abnormalities, U-net is widely used \autocite{Ronneberger_Fischer_Brox_2015}. With regards to training CNNs, most studies use transferred learning, which consists of three steps: learning, fine-tuning, and validation.

	Some studies directly predict the disease from OCT images.
	\authyear{li2019deep} used ResNet to analyze OCT images and distinguish between choroidal neovascularization (CNV), diabetic macular edema (DME), drusen, and healthy eyes. In addition, they performed occlusion testing to find out the regions that are the most important in diagnosis \autocite{li2019deep}. 
	\authyear{yoo2021feasibility} used generative adversarial network (GAN) and Inception-v3 together with a few-shot dataset to investigate the feasibility of improving OCT diagnosis of rare ocular diseases \autocite{yoo2021feasibility}. They handled the issues of training image shortage and data imbalance by creating ocular disease OCT images from healthy OCT images. \authyear{Kermany2018} conducts medical diagnosis and identifies treatable diseases by image-based deep learning. They also provided a widely used database including more than one hundred thousand OCT labeled images for 3 diseases \autocite{Kermany2018}.

	Some other studies can identify ocular abnormalities from annotated OCT images.
	\authyear{camino2018deep} used DL to identify the region of preserved photoreceptors on \textit{en face} OCT in choroideremia and retinitis pigmentosa (RP) \autocite{camino2018deep}. 
	\authyear{srinivasan2014fully} detected DME and dry age-related macular degeneration (dry AMD) from OCT images by flattening the image and using support vector machine (SVM) to extract the thickness information of the retinal layers \autocite{srinivasan2014fully}. 
	\authyear{leandro2023oct} implemented VGG to identify up to 8 kinds of key abnormalities and hence detect multiple diseases by using central fovea cross-section OCT \autocite{leandro2023oct}. \authyear{Fang_Wang2019} developed a novel lesion-aware CNN, called LACNN, to simulate the ophthalmologists' diagnosis that focuses on ocular abnormalities. The LACNN is a U-net-like CNN which incorporates VGG16 as the baseline, and the result is impressive \autocite{Fang_Wang2019}.

	Other studies use fundus images to predict ocular diseases.
	\authyear{masumoto2019accuracy} trained deep CNN with ultrawide-field fundus images to make the CNN capable of diagnosing RP \autocite{masumoto2019accuracy}. 
	\authyear{chen2021artificial} uses color fundus photographs and multiple types of CNN (Inception V3, Inception Resnet V2, and Xception) to develop a method of early detection of RP \autocite{chen2021artificial}. 
	\authyear{li2022development} uses CNN to detect up to 12 fundus diseases based on colour fundus photography \autocite{li2022development}. \authyear{Son2023} presented a novel architectural and algorithmic design to comprehensively identify 15 abnormalities and diagnose 8 major ocular diseases from macula-centered fundus images. They defined a notion of counterfactual attribution ratio (CAR) to interpret the system's diagnostic reason and disclose the relationship between abnormalities and diseases \autocite{Son2023}.

	In recent years, more and more DL systems began to use multi-modal information to predict ocular diseases. For automated detection system, retrieving features from both OCT images and fundus images can effectively keep the diagnosis away from bias and incompleteness. It is not always necessary to yield a better automated diagnosis result, but it does help ophthalmologist to make more accurate and holistic clinical decisions. For instance, \authyear{liu2023prediction} combined OCT and fundus images to evaluate visual impairment in RP in terms of best-corrected visual acuity(BCVA) \autocite{liu2023prediction}. \authyear{Xu2021} leveraged a bi-modal CNN to diagnose AMD and polypoidal choroidal vasculopathy (PCV), where the architecture uses fundus and OCT images as input for transferred learning CNN and concatenates the retrieved features to classify 3 diseases including dry AMD, wet AMD and PCV \autocite{Xu2021}. And \author{Andrearczyk2018} ingeniously combined medical images and biomedical textual information to discriminate different diseases, which is a novel method for the usage of multi-modal application \autocite{Andrearczyk2018}.

	The key technical problem of multi-modal diagnosis is how to fuse the results from multifarious sources. There are three types of fusion, namely early, late, and hybrid fusion. Deciding on the optimal type of fusion is part of the exploratory process in the application of DL methods \autocite{Ichhpujani_Thakur_2021}.

	\section{Overview}
	\label{sec:overview}

	We develop the Multimodal Abnormity-Aware Model (MAAM) to aid ophthalmologists in diagnosing ocular diseases. It considers both OCT and color fundus image inputs. In addition, it simulates the decision-making process of ophthalmologists when they examine patients, namely, first identifying the abnormities, and then deducing the disease. The model is designed in a way that partly opens up the ``black box'' of DL and presents more interpretable information to ophthalmologists for the reference.

	The structure of the MAAM is shown in Fig.~\ref{fig:3_parts}. It consists of 2 submodels: the Abnormity Model and the Diagnosis Model. We input multiple OCT and fundus images to the Abnormity Models, and output the classification of abnormities as an input to the Diagnosis Model, which yields disease probabilities as the final result.

@@

	The Abnormity Models are de facto classifiers for abnormities, which are symptoms that can be directly observed in the images. The MAAM identifies 11 OCT abnormities and 8 Fundus abnormities, as shown in Fig.~\ref{fig:OCT_abnormities} and Fig.~\ref{fig:fundus_abnormities}. The abbreviations and detail information of abnormities and diseases are tabulated in ``Appendix''.

@@

	\begin{figure}[htbp]
	\centering
	\includegraphics[width=\linewidth]{Figs/model_overview.pdf}
	\caption{Model overview}
	\vspace{0.3cm}
	\label{fig:3_parts}
	\end{figure}

	\begin{figure}[htbp]
		\centering
		\includegraphics[width=\linewidth]{Figs/OCT_Abnormities.pdf}
		\caption{OCT abnormities \autocite{Duker_Waheed_Goldman_2022}}
		\vspace{0.3cm}
		\label{fig:OCT_abnormities}
	\end{figure}

	\begin{figure}[htbp]
		\centering
		\includegraphics[width=\linewidth]{Figs/fundus_Abnormities.pdf}
		\caption{Fundus abnormities \autocite{Wolf_Kirchhof_Reim_2006}}
		\vspace{0.3cm}
		\label{fig:fundus_abnormities}
	\end{figure}

@@

	The Abnormity Models also contain 2 submodels: the OCT Model and the Fundus Model, which classify OCT and fundus abnormities, respectively. Both of the submodels leverage CNNs with modified final fully-connected (FC) layers. We compare the performances of 4 commonly used CNNs, ResNet152, ResNet50, ResNet18 \autocite{He_Zhang_Ren_Sun_2016} and VGG16 \autocite{Simonyan_Zisserman_2015}, and choose the best one for each submodel. The final FC layer in each CNN is modified so that the size of its output vector is equal to the number of OCT or fundus abnormities. In order to normalize the output vector, we perform the softmax operation. Each submodel outputs a probability vector for all the abnormities.

	The Diagnosis Model consists of two stages: Stage D1 and Stage D2.

@@

	In Stage D1, we determine the severity level for each disease from the probability vectors from the Abnormity Models. We use Abnormity-to-Disease Deduction Criteria (shown in Fig.~\ref{fig:criteria}) as ground truth. There are multiple submodels in Stage D1 and each one corresponds to one disease. Based on the deduction criteria, we determine the number of abnormities for one disease and use the number to define the severity levels. For example, there are totally 5 abnormities that are present in the disease ``rDR'': OCT abnormity ``DME'' and Fundus abnormities ``HM'', ``VA'', ``MA'', and ``CWP''. Therefore, taking into account the healthy status, we end up with 6 severity levels for disease ``rDR''.

	We use the fused vector as an input for each submodels in Stage D1, and have the vector go through a FC layer with softmax operation to yield severity level probability vectors.

	\begin{figure}[htbp]
		\centering
		\includegraphics[width=0.7\linewidth]{Figs/criteria.pdf}
		\caption{Abnormity-to-disease Deduction Criteria}
		\vspace{0.3cm}
		\label{fig:criteria}
	\end{figure}

	\begin{multicols}{2}	
	In Stage D2, we determine the final disease probability vector. Similarly, we use a fused vector from the outputs of submodels in Stage D1, and have the vector go through a FC layer with softmax operation. The result presents potential diseases that can be referenced by ophthalmologists.

	In the MAAM, we leverage the fusion operation to incorporate all the results from different submodels. As shown in Fig.~\ref{fig:fusion}, there are totally 3 fusion operations in MAAM.

	The first fusion occurs at the output of OCT Model or Fundus Model. In real scenarios, multiple OCT and Fundus images can be used for ocular disease diagnosis. Each image goes through either the OCT Model or the Fundus Model to yield a probability vector. To use the probability results from all the images, a maximization fusion is implemented for all the vectors so that the fusion yields one OCT abnormity probability vector and one Fundus abnormity probability vector with maximum values.

	The second fusion occurs at the interface between Abnormity Model and Stage D1. A concatenation fusion is implemented for OCT abnormity probability vector and Fundus abnormity probability vector. The fused vector works as an input to all the submodels in Stage D1.

	The third fusion occurs at the interface between Stage D1 and Stage D2. We concatenation all the severity level vectors to feed the model in Stage D2. The fused vector goes through the FC layer and undergoes a softmax operation to yield the final result.

	\begin{figure}[htbp]
	\centering
	\includegraphics[width=\linewidth]{Figs/fusion.pdf}
	\caption{Fusion mechanism}
	\vspace{0.3cm}
	\label{fig:fusion}
	\end{figure}

	\section{Abnormity Models}

	\subsection{Data Preparation}

	The images and labels used for training are mainly downloaded from public databases, but some abnormities are not included in these databases. For those abnormities, we use the search engine as an additional data source to get images. The numbers of images acquired from each data source are shown in Table~\ref{tb:OCT_source} and Table~\ref{tb:Fundus_source}. As to the links of the images from the search engine, refer to ``Appendix''.

	For OCT abnormities with less than 1000 images, we use Cycle-GAN \autocite{Zhu_Park_Isola_Efros_2020} to generate new images. We train a Cycle-GAN network to interconvert healthy and abnormity images, take out the generator that converts healthy to abnormity images, and use it to generate new abnormity images. We go through all generated images and adopt those that clearly display only the abnormity of interest. The adoption rates are shown in Table~\ref{tb:cycleGAN_number}. Fundus images are not generated using Cycle-GAN because generated fundus images are too blurry; in addition, fundus abnormities are usually less salient than OCT ones and cannot be properly learned by Cycle-GAN, given that we only have a small number of fundus images.
	\end{multicols}

	\vspace{-0.5cm}
	\begin{minipage}[t]{0.4\linewidth}
		{
			\fontsize{9}{12}\selectfont
			{
				\begin{longtable}{cccccc}
					\caption{OCT Data sources}
					\label{tb:OCT_source}\\
					\toprule
					\multirow{2}{*}{Abnormity}&\multicolumn{4}{c}{Source}&\multirow{2}{*}{Total}\\
					&1&2&3&4&\\
					\midrule
					CNV    &2984&-  &- &- &2984\\
					CSC    &-   &102&32&- &134 \\
					DME    &2500&-  &- &- &2500\\
					Drusen &2500&-  &- &- &2500\\
					ERM    &-   &-  &- &16&16  \\
					IF     &1097&-  &- &- &1097\\
					MH     &-   &99 &31&- &130 \\
					Mactel &-   &-  &29&- &29  \\
					Healthy&5000&-  &- &- &5000\\
					RP     &-   &102&31&- &133 \\
					ST     &-   &-  &23&- &23  \\
					SF     &1083&-  &- &- &1083\\
					\bottomrule
				\end{longtable}

@@

					\item Search engine
					\vspace{-0.2cm}
				\end{enumerate}

				\vspace{0.5cm}
			}
		}
	\end{minipage}
	\begin{minipage}[t]{0.6\linewidth}
		{
			\fontsize{9}{12}\selectfont
			{
				\begin{longtable}{cccccccc}
					\caption{Fundus Data sources}
					\label{tb:Fundus_source}\\
					\toprule
					\multirow{2}{*}{Abnormity}&\multicolumn{6}{c}{Source}&\multirow{2}{*}{Total}\\
					&1&2&3&4&5&6&\\
					\midrule
					CWP    &-  &- &- &33 &205&- &238\\
					Drusen &-  &- &- &50 &-  &- &50 \\     
					HE     &20 &- &- &75 &284&- &379\\ 
					HM     &13 &66&- &105&278&- &462\\     
					MH     &-  &- &- &-  &-  &34&34 \\        
					MA     &55 &- &- &1  &219&- &275\\
					Healthy&100&37&15&-  &-  &- &152\\      
					RP     &-  &22&- &-  &-  &44&66 \\         
					VA     &-  &64&- &14 &-  &- &78 \\

@@

					\item Search engine
					\vspace{-0.2cm}

@@

	{
		\fontsize{9}{12}\selectfont
		{
			\begin{longtable}{ccccc}
				\caption{Adoption rates of Cycle-GAN-generated OCT images}
				\label{tb:cycleGAN_number}\\
				\toprule
				Abnormity&Original&Generated&Adopted&Adoption Rate\\
				\midrule
				CSC   &134&3000&1730&58.667\% \\
				ERM   &19 &3000&1931&64.367\% \\
				MH    &130&3000&1811&60.367\% \\
				Mactel&29 &3000&1900&63.333\% \\
				RP    &133&3000&1923&63.100\% \\
				ST    &23 &3000&2572&86.067\% \\
				\bottomrule
			\end{longtable}
		}
	}

	\begin{multicols}{2}
	We split the images, including generated images, into training and test datasets, and implement a set of transformations on the images, including horizontal flips, random brightness changes from -10\% to +10\%, random horizontal and vertical translations between -5\% and +5\%, random scaling between -20\% and +20\%, and random rotations, which are between -10$^\circ$ and +10$^\circ$ for OCT images and between -30$^\circ$ and +30$^\circ$ for fundus images. The rotation is only applied to images in the train dataset. Eventually, we get 5000 train images and 500 test images for each OCT abnormity, and 3000 train images and 300 test images for each fundus abnormity.

	\subsection{Training}
	\label{sec:a_training}

	The model training is on a desktop computer with Intel$^®$ Xeon$^®$ Platinum 8352V Processor, 256GB of RAM and 2 NVIDIA GPU (GeForce RTX 4090) with 48GB VRAM. The training uses cross entropy loss, ADAM optimizer with learning rate 0.001, a batch size of 32 and five-fold cross-validation. The code is written with PyTorch in an Anaconda environment. Refer to ``Appendix'' for the link to the code repository on GitHub.

@@

	For the Abnormity Models, we employ a strategy of transfer learning to finetuning.

>>>>

\documentclass{article}
\usepackage{ctex}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{amsmath}
\usepackage{caption}
\RequirePackage{longtable,multirow,array}
\RequirePackage{booktabs}
\usepackage{booktabs}
\usepackage{multicol}
\usepackage{longtable}
\usepackage{xurl}
\usepackage{booktabs}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{enumitem}
\usepackage[style=ieee,sorting=nyt]{biblatex}
\DefineBibliographyStrings{english}{
	references = {参考文献}
}
\assignrefcontextentries[]{*}
\addbibresource{References.bib}
\geometry{left=1in, right=1in, top=1in, bottom=1in}
\hypersetup{hidelinks}
\hypersetup{bookmarksnumbered=true,
	bookmarksopen=true,
	colorlinks=true,
	linkcolor=black,
	citecolor=blue,
	urlcolor=black}
\usepackage{tikz}
\usepackage{pgfplotstable}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\captionsetup{justification=centering}
\captionsetup[figure]{name={图},labelsep=space,labelfont=bf}
\captionsetup[table]{name={表},labelsep=space,labelfont=bf}
\CTEXoptions[today=old, contentsname=Contents]
\newcommand{\authyear}[1]{\citeauthor{#1} (\citeyear{#1})}
\setmainfont{Arial}

@@

\begin{document}
	\title{\vspace{-2.25cm} 眼科疾病诊断的多模态异常感知模型}
	\author{潘思齐}
	\date{}
	\maketitle

	\vspace{-0.2cm}
	\hrule
	\vspace{-0.1cm}
	\section*{摘要}
	\hrule
	\vspace{0.4cm}
	我们设计了一种新颖的模型，即多模态异常感知模型（MAAM），用于辅助眼科疾病的诊断。该模型允许输入两类眼科图像，分别是光学相干断层扫描（OCT）图像和彩色眼底图像，并基于异常来推导疾病，模拟眼科医生的决策过程。该模型由两个子模型组成，分别是异常模型和诊断模型。异常模型识别11种OCT异常和8种眼底异常；诊断模型通过整合异常模型的特征识别出12种疾病。在MAAM中，我们引入了融合机制、共显异常和疾病的概念、以及异常-疾病推导标准。该模型打破了复杂神经网络的“黑箱”，为眼科医生提供了更具可解释性的信息以供参考。模型的总体准确率为89\%，验证了其可行性。

	\vspace{0.4cm}
	\hrule
	\vspace{1cm}

	\section{背景}

	眼科疾病可以通过多种方法诊断，包括使用光学相干断层扫描（OCT）和彩色眼底图像。眼科医生通常通过识别眼部异常来推导疾病。传统上，诊断主要依赖于眼科医生的专业经验和知识，这可能导致高误诊率和医疗数据的利用不足。随着人工智能（AI）的广泛应用，深度学习（DL）共享专家知识，为偏远地区的患者提供支持方面做出了巨大贡献 \autocite{Ichhpujani_Thakur_2021}。通过利用深度学习，研究人员开发了辅助诊断程序，以帮助眼科医生进行诊断。许多研究都使用卷积神经网络（CNN）来分析眼科图像。一些常用的 CNN 包括 VGG、ResNet 和 Inception \autocite{daich2023artificial}。在图像分割和寻找异常中，U-net 被广泛应用 \autocite{Ronneberger_Fischer_Brox_2015}。大多数研究在训练CNN时使用迁移学习，包括三个步骤：学习、微调和验证。

	一些研究直接从OCT图像预测疾病。\citeauthor{li2019deep} 使用ResNet分析OCT图像，并区分脉络膜新生血管（CNV）、糖尿病性黄斑水肿（DME）、玻璃疣（Drusen）和健康人眼。此外，他们还进行了遮挡测试，以找出对于诊断最重要的区域 \autocite{li2019deep}。\citeauthor{yoo2021feasibility} 使用生成对抗网络（GAN）和Inception-v3，结合少量数据集，探讨了改善罕见眼科疾病OCT诊断的可行性 \autocite{yoo2021feasibility}。他们通过从健康OCT图像中生成眼科疾病的OCT图像来解决图像数量不足和数据不平衡的问题。\citeauthor{Kermany2018} 通过基于图像的深度学习进行医学诊断，并识别可治疗的疾病。他们还提供了一个广为使用的数据库，其中包含超过十万张用于三种疾病的OCT的有标注图像 \autocite{Kermany2018}。

	其他研究可以从带注释的OCT图像中识别眼科异常。\citeauthor{camino2018deep} 利用深度学习在面部OCT中识别未凋亡的的光感受器区域，用于脉络膜疾病和视网膜色素变性（RP） \autocite{camino2018deep}。\citeauthor{srinivasan2014fully} 通过展平图像，使用支持向量机（SVM）提取视网膜层的厚度信息，检测 DME 和干性年龄相关性黄斑变性（干性AMD） \autocite{srinivasan2014fully}。\citeauthor{leandro2023oct} 使用VGG，用中央黄斑横截面OCT检测多达8种关键异常并由此检测多种疾病 \autocite{leandro2023oct}。\citeauthor{Fang_Wang2019} 开发了一种新的病灶感知 CNN，称为 LACNN，用于模拟集中于眼科异常的眼科医生诊断 \autocite{Fang_Wang2019}。

	另一些研究使用眼底图像来预测眼科疾病。\citeauthor{masumoto2019accuracy} 使用超宽视野眼底图像训练深度 CNN 以诊断 RP \autocite{masumoto2019accuracy}。\citeauthor{chen2021artificial} 使用彩色眼底照片和多种 CNN（如 Inception V3、Inception Resnet V2 和 Xception），开发了 RP 的早期检测方法 \autocite{chen2021artificial}。\citeauthor{li2022development} 使用 CNN 基于彩色眼底摄影检测多达 12 种眼底疾病 \autocite{li2022development}。\citeauthor{Son2023} 提出了一种新颖的架构和算法设计，从黄斑中心的眼底图像中综合识别 15 种异常和诊断 8 种主要眼科疾病。他们定义了反事实归因比（CAR）的概念，以解释系统的诊断原因，并揭示异常与疾病之间的关系 \autocite{Son2023}。

	近年来，越来越多的深度学习系统开始使用多模态信息来预测眼科疾病。对于自动化检测系统，从 OCT 图像和眼底图像中提取特征可以有效避免诊断偏差和不完整性。虽然并不总是能带来更好的自动诊断结果，但它确实有助于眼科医生做出更准确和全面的临床决策。例如，\citeauthor{liu2023prediction} 结合 OCT 和眼底图像，评估视网膜色素变性患者的视力损害 \autocite{liu2023prediction}。\citeauthor{Xu2021} 利用双模态 CNN 诊断 AMD 和脉络膜息肉样血管病变（PCV），其中架构使用眼底和 OCT 图像作为迁移学习 CNN 的输入，并将提取的特征拼接以分类干性 AMD、湿性 AMD 和 PCV 三种疾病 \autocite{Xu2021}。\citeauthor{Andrearczyk2018} 将医学图像与生物医学文本信息结合起来，以区分不同的疾病，这是多模态应用的一种创新方法 \autocite{Andrearczyk2018}。

	多模态诊断的关键技术问题在于如何融合来自多种来源的结果。融合有三种类型，即早期融合、后期融合和混合融合。决定最佳融合类型是应用深度学习方法探索过程的一部分 \autocite{Ichhpujani_Thakur_2021}。

	\section{概述}
	\label{sec:overview}

	我们开发了多模态异常感知模型（MAAM）来协助眼科医生诊断眼科疾病。该模型同时考虑 OCT 和彩色眼底图像输入。此外，它模拟了眼科医生在检查患者时的决策过程，即首先识别异常，然后推导疾病。该模型设计部分打开了深度学习的“黑箱”，并为眼科医生提供了更多可解释的信息参考。

	MAAM 的结构如图~\ref{fig:3_parts} 所示。模型由两个子模型组成：异常模型和诊断模型。我们将多个 OCT 和眼底图像输入到异常模型中，并将异常分类结果作为诊断模型的输入，诊断模型最终输出疾病概率。

@@

	异常模型实际上是异常分类器，即图像中可直接观察到的症状。MAAM 能识别 11 种 OCT 异常和 8 种眼底异常，如图~\ref{fig:OCT_abnormities} 和图~\ref{fig:fundus_abnormities} 所示。“附录”中列出了异常和疾病的缩写及详细信息。

	异常模型还包括两个子模型：OCT 模型和眼底模型，分别对 OCT 和眼底异常进行分类。两个子模型都使用 CNN，并修改了最终的全连接层（FC 层）。我们比较了四种常用 CNN 的性能：ResNet152、ResNet50、ResNet18 \autocite{He_Zhang_Ren_Sun_2016}和 VGG16 \autocite{Simonyan_Zisserman_2015}，并为每个子模型选择了最佳模型。每个 CNN 的最终 FC 层被修改，使其输出向量的大小等于 OCT 或眼底异常的数量。为了对输出向量进行归一化，我们对其进行了 softmax 操作。每个子模型输出所有异常的概率向量。

	诊断模型包括两个阶段：阶段 D1 和阶段 D2。

	在阶段 D1 中，我们根据异常模型的概率向量确定每种疾病的严重程度。我们使用异常到疾病推导标准（见图~\ref{fig:criteria}）作为判断依据。阶段 D1 中有多个子模型，每个子模型对应一种疾病。根据推导标准，我们确定一种疾病的异常数量，并以此来定义严重程度。例如，疾病“rDR”中共有 5 种异常：OCT 异常“DME”以及眼底异常“HM”、“VA”、“MA”和“CWP”。因此，考虑到健康状态，我们将“rDR”定义为 6 个严重程度级别。

	我们使用融合后的向量作为阶段 D1 中每个子模型的输入，让向量通过一个 FC 层和 softmax 操作，以生成严重程度的概率向量。

	在阶段 D2 中，我们确定最终的疾病概率向量。同样，我们使用阶段 D1 中子模型的输出，进行向量融合，通过 FC 层并进行 softmax 操作，输出潜在疾病的概率，供眼科医生参考。

@@

	\begin{figure}[htbp]
		\centering
		\includegraphics[width=\linewidth]{Figs/model_overview.png}
		\caption{模型总览}
		\vspace{0.3cm}
		\label{fig:3_parts}
		\vspace{1cm}
	\end{figure}

@@

	在 MAAM 中，我们利用融合操作整合不同子模型的结果。如图~\ref{fig:fusion} 所示，MAAM 中总共有三次融合操作。

	第一次融合发生在 OCT 模型或眼底模型的输出中。在实际场景中，可以使用多个 OCT 和眼底图像进行眼科疾病诊断。每张图像通过 OCT 或眼底模型后得到一个概率向量。为了利用所有图像的概率结果，我们对所有向量进行了最大化融合，从而得到一个 OCT 异常概率向量和一个眼底异常概率向量，均取最大值。

	第二次融合发生在异常模型与阶段 D1 之间的接口。我们将 OCT 异常概率向量和眼底异常概率向量进行拼接融合，作为阶段 D1 所有子模型的输入。

	第三次融合发生在阶段 D1 与阶段 D2 之间的接口。我们将所有严重程度向量进行拼接，输入阶段 D2 模型。融合向量通过 FC 层并经过 softmax 操作，输出最终结果。

	\section{异常模型}

	\subsection{数据准备}

	用于训练的图像和标签主要从公共数据库中下载，但部分异常不包含在这些数据库中。对于这些异常，我们使用搜索引擎作为额外的数据来源获取图像。各数据来源获得的图像数量如表~\ref{tb:OCT_source} 和表~\ref{tb:Fundus_source} 所示。图像的搜索引擎链接详见“附录”。

	对于图像少于 1000 张的 OCT 异常，我们使用 Cycle-GAN \autocite{Zhu_Park_Isola_Efros_2020}生成新图像。我们训练一个 Cycle-GAN 网络，将健康图像与异常图像进行转换，取出将健康图像转换为异常图像的生成器，用其生成新的异常图像。我们检查所有生成的图像，选取那些清晰显示了目标异常的图像。采纳率见表~\ref{tb:cycleGAN_number}。由于生成的眼底图像过于模糊，眼底异常通常不如 OCT 明显，Cycle-GAN 无法从少量眼底图像中学习，因此我们不使用 Cycle-GAN 生成眼底图像。

	\begin{figure}[htbp]
		\centering
		\includegraphics[width=\linewidth]{Figs/OCT_Abnormities.pdf}
		\caption{OCT异常 \autocite{Duker_Waheed_Goldman_2022}}
		\vspace{0.3cm}
		\label{fig:OCT_abnormities}
	\end{figure}

	\begin{figure}[htbp]
		\centering
		\includegraphics[width=\linewidth]{Figs/fundus_Abnormities.pdf}
		\caption{眼底异常 \autocite{Wolf_Kirchhof_Reim_2006}}
		\vspace{0.3cm}
		\label{fig:fundus_abnormities}
	\end{figure}

	\begin{figure}[htbp]
		\centering
		\includegraphics[width=0.7\linewidth]{Figs/criteria.png}
		\caption{异常-疾病推导标准}
		\vspace{0.3cm}
		\label{fig:criteria}
	\end{figure}

@@

	\begin{figure}[htbp]
	\centering
	\includegraphics[width=\linewidth]{Figs/fusion.png}
	\caption{融合机制}
	\vspace{0.3cm}
	\label{fig:fusion}
	\end{figure}

	\begin{multicols}{2}
	我们将包括生成图像在内的图像分成训练和测试数据集，并对图像进行一系列变换，包括水平翻转、亮度随机变化（-10\% 到 +10\%）、水平和垂直随机平移（-5\% 到 +5\%）、随机缩放（-20\% 到 +20\%）和随机旋转（OCT 图像为 -10° 到 +10°，眼底图像为 -30° 到 +30°）。旋转仅应用于训练数据集中的图像。最终，我们为每个 OCT 异常获得 5000 张训练图像和 500 张测试图像，为每个眼底异常获得 3000 张训练图像和 300 张测试图像。
	\end{multicols}

		\vspace{-0.5cm}
		\begin{minipage}[t]{0.4\linewidth}
			{
				\fontsize{9}{12}\selectfont
				{
					\begin{longtable}{cccccc}
						\caption{OCT数据源}
						\label{tb:OCT_source}\\
						\toprule
						\multirow{2}{*}{异常}&\multicolumn{4}{c}{数据源}&\multirow{2}{*}{总数据量}\\
						&1&2&3&4&\\
						\midrule
						CNV    &2984&-  &- &- &2984\\
						CSC    &-   &102&32&- &134 \\
						DME    &2500&-  &- &- &2500\\
						Drusen &2500&-  &- &- &2500\\
						ERM    &-   &-  &- &16&16  \\
						IF     &1097&-  &- &- &1097\\
						MH     &-   &99 &31&- &130 \\
						Mactel &-   &-  &29&- &29  \\
						Healthy&5000&-  &- &- &5000\\
						RP     &-   &102&31&- &133 \\
						ST     &-   &-  &23&- &23  \\
						SF     &1083&-  &- &- &1083\\
						\bottomrule
					\end{longtable}

@@

						\item 搜索引擎
						\vspace{-0.2cm}
					\end{enumerate}

					\vspace{0.5cm}
				}
			}
		\end{minipage}
		\begin{minipage}[t]{0.6\linewidth}
			{
				\fontsize{9}{12}\selectfont
				{
					\begin{longtable}{cccccccc}
						\caption{眼底数据源}
						\label{tb:Fundus_source}\\
						\toprule
						\multirow{2}{*}{异常}&\multicolumn{6}{c}{数据源}&\multirow{2}{*}{总数据量}\\
						&1&2&3&4&5&6&\\
						\midrule
						CWP    &-  &- &- &33 &205&- &238\\
						Drusen &-  &- &- &50 &-  &- &50 \\     
						HE     &20 &- &- &75 &284&- &379\\ 
						HM     &13 &66&- &105&278&- &462\\     
						MH     &-  &- &- &-  &-  &34&34 \\        
						MA     &55 &- &- &1  &219&- &275\\
						Healthy&100&37&15&-  &-  &- &152\\      
						RP     &-  &22&- &-  &-  &44&66 \\         
						VA     &-  &64&- &14 &-  &- &78 \\

@@

						\item 搜索引擎
						\vspace{-0.2cm}

@@

		{
			\fontsize{9}{12}\selectfont
			{
				\begin{longtable}{ccccc}
					\caption{Cycle-GAN生成的OCT图片的采用率}
					\label{tb:cycleGAN_number}\\
					\toprule
					异常&原图数量&生成数量&采用数量&采用率\\
					\midrule
					CSC   &134&3000&1730&58.667\% \\
					ERM   &19 &3000&1931&64.367\% \\
					MH    &130&3000&1811&60.367\% \\
					Mactel&29 &3000&1900&63.333\% \\
					RP    &133&3000&1923&63.100\% \\
					ST    &23 &3000&2572&86.067\% \\
					\bottomrule
				\end{longtable}
			}
		}

	\begin{multicols}{2}
	\subsection{训练}
	\label{sec:a_training}

		模型训练在一台台式计算机上进行，配备 Intel$^®$ Xeon$^®$ Platinum 8352V 处理器、256GB 内存和 2 张 NVIDIA GeForce RTX 4090 GPU（48GB 显存）。训练使用交叉熵损失、学习率为 0.001 的 ADAM 优化器、批大小为 32 以及五折交叉验证。代码在 Anaconda 环境下用 PyTorch 编写。代码库的 GitHub 链接参见“附录”。

@@

		对于异常模型，我们采用迁移学习微调策略。

		在迁移学习阶段，使用 ImageNet \autocite{Krizhevsky_Sutskever_Hinton_2017}的预训练权重。我们冻结所有卷积层中的权重，仅调整最终 FC 层的权重。训练持续 100 个 epoch。每 10 个 epoch，保存验证集准确率最高的模型权重。为防止训练数据集的过拟合，我们引入总体准确率，计算验证集和测试集的加权平均准确率。对于所有保存的模型，计算其总体准确率，从中选择验证集和测试集表现最好的模型，进行微调。

		在微调阶段，从迁移学习阶段的最佳模型开始，解冻模型中的所有权重。训练 30 个 epoch。每 10 个 epoch 保存验证集准确率最高的模型权重，选出总体准确率最高的模型。最终比较迁移学习阶段和微调阶段的最佳模型的总体准确率，以确定最佳模型。
